Transient Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-29 paper/ms.tex
[WIP] Update ms with JNW suggestions (intro+abs)
Co-authored-by: Joshua Winn <jnwinn@princeton.edu>

All pre-main-sequence stars vary in optical brightness, and the origin
of such variability is, in most cases, understood.  Well-explored
sources of optical variability include inhomogeneities on stellar
surfaces such as starspots and faculae
\citep[e.g.][]{2021isma.book.....B}, occultations by gas-rich
circumstellar disks \citep[e.g.][]{2017MNRAS.470..202B}, and, in
geometrically favorable circumstances, eclipses by stars and planets
\citep[e.g.][]{2010exop.book...55W}.  More exotic forms of optical
variability relevant to this work include transiting exocomets
(e.g.~$\beta$~Pic; \citealt{2019A&A...625L..13Z}) and disintegrating
rocky bodies around both M dwarfs (e.g. KOI-2700;
\citealt{2014ApJ...784...40R}) and white dwarfs (e.g. WD~1145;
\citealt{2015Natur.526..546V}).

Data from K2 and TESS have yielded a new class of variable star whose
root cause is only beginning to become clear: complex quasiperiodic
variables (CQVs).  These objects are identified from their optical
light curves, which show nearly periodic troughs that are either sharp
or broad; these troughs are often superposed on smooth spot-like
modulation
\citep{2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z}.
Some CQVs show up to eight local minima (``dips'') per cycle.  Most
are pre-main-sequence M dwarfs without near-infrared excesses, ages of
$\approx$5-150 million years (Myr), and rotation periods of at most
two days; they are observed to comprise $\approx$1-3\% of M dwarfs
younger than 100 Myr \citep{2016AJ....152..114R,2022AJ....163..144G}.
The dips can be chromatic, with a reddening law plausibly consistent
with dust
\citep{2017PASJ...69L...2O,2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
And finally, while the dip shapes can ``jump'' between different
depths and durations over less than one cycle, they more often evolve
gradually, over tens to hundreds of cycles
\citep[e.g.][]{2017AJ....153..152S,2022ApJ...925...75P,2023ApJ...945..114P}.

The CQV light curves cannot be explained by pure starspots.  Although
all young M dwarfs are spotted, this tends to produce flux variations
over timescales of the rotation period, $P_{\rm rot}$,
and its half-harmonic, $0.5\,P_{\rm rot}$.  With fine-tuned
viewing geometries, starspots can produce dip durations as short
as $\approx$$0.2\,P_{\rm rot}$, and the dip amplitudes in such
geometries are small
due to limb-darkening (see \citealt{2017AJ....153..152S},
Figures 37-41).  The observed dips occur over durations as
short as $0.05\,P_{\rm rot}$ with characteristic depths of a few
percent; a ``starspot-only'' scenario can be discarded for any object
with dips that are sufficiently large in depth and short in duration
\citep{2017AJ....153..152S,2019ApJ...876..127Z,2021MNRAS.500.1366K}.   The dip durations and
amplitudes instead require
sharp geometries with material extrinsic to the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  Since some CQVs require
such geometries, and the
CQVs probably all have a single explanation, we therefore
discard the ``spot-only'' scenario.

A few competing explanations for what causes the complex quasiperiodic
variability are shown in Figure~\ref{fig:f1}.  The ``clump'' scenario
invokes opaque dust clumps periodically transiting the star near the
Keplerian corotation radius, $R_{\rm cr} = (GM/\Omega^2)^{1/3}$
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The ``prominence''
scenario invokes long-lived condensations of cool and dense
marginally-ionized gas, embedded within the hotter corona, that would
be centrifugally supported to also be near corotation
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
Such structures are analogous to quiescent prominences and filaments
seen in the solar corona \citep[see e.g.][]{2015ASSL..415.....V},
though at much larger relative distances from the stellar surface.  A
final possibility is one of a ``screen'', in which the inner wall of a
quiescent circumstellar disk blocks a portion of the stellar surface
to produce sudden dips whenever spots come into view
\citep{2019ApJ...876..127Z}.

The dust clump and prominence hypotheses seem most plausible.  They
are qualitatively similar, except that one invokes opacity from dust,
while the other invokes opacity from gas, likely bound-free
transitions in hydrogen or perhaps a molecular opacity.  The screen
scenario seems inconsistent with the sharpness of the observed dips,
the lack of infrared excess, and the observed lifetime of CQVs
extending an order of magnitude longer than the $\approx$$10^7$ year
timescale often quoted for primordial disk dispersal.  However,
unambiguous evidence in support of any one scenario has yet to be
acquired.  Such evidence might include a spectroscopic detection of
silicate 10\,$\mu$m dust absorption during a dip, or perhaps detection
of transient Balmer-line excesses as a function of cycle phase,
similar to observations made in systems such as AB~Dor
\citep[see][]{1999ASPC..158..146C} or PTFO 8-8695
\citep{2016ApJ...830...15J}.

CQVs remain mysterious because they have been both hard to discover
and hard to characterize.   Discoverability is tied to rarity: CQVs
comprise $\approx$1\% of the youngest $\approx$1\% of M dwarfs
\citep{2018AJ....155..196R}.  Out of the millions of stars monitored
by K2 and TESS, about 50 CQVs have been reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2022AJ....163..144G,2023ApJ...945..114P}.
The known CQVs are correspondingly faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, yielding optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  This renders time-series spectroscopy
at high resolution out of reach for current facilities, despite its
potential utility in ruling out some of these models.

In this work, we aim to find bright and nearby CQVs, since these
objects will be the most amenable to detailed photometric and
spectroscopic analyses.  To do this, we use 120-second cadence data
acquired by TESS between 2018 July and 2022 September (Sectors 1-55;
Cycles 1-4).  We present our search methods in
Section~\ref{sec:methods}, and the properties of the resulting CQV
catalog in Section~\ref{sec:results}.  The evolution of many CQVs over
a two-year baseline is described in Section~\ref{sec:evoln}, including
a deep-dive into LP 12-502.  We discuss a few implications in
Section~\ref{sec:discussion}, and conclude in
Section~\ref{sec:conclusion}.

A point on nomenclature.  CQVs have been called ``transient flux
dips'', ``persistent flux dips'', ``scallop shells'', ``batwings'',
\citep{2017AJ....153..152S} ``complex rotators'',
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P},
and ``complex periodic variables'' \citep{2023MNRAS.518.2921K}.  The
CQVs should not be conflated with ``dippers'', which are classical T
Tauri stars with infrared excesses, and which show large-amplitude
variability linked to obscuring inner disk structures and accretion
hot spots \citep{2014AJ....147...82C,2021ApJ...908...16R}.  While a
few similarities between CQVs and dippers do exist (see
Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}), their
phenomenology and stellar properties are quite different.  At the risk
of introducing yet another standard, we hope to use a nomenclature
that reflects how, when observed over timescales of more than tens of
cycles, CQVs are almost but not exactly periodic.  They are
quasiperiodic.  While the three-type classification scheme proposed by
\citet{2017AJ....153..152S} may provide helpful visual
distinctions amongst the CQVs, it seems likely that they are all
explained by a single underlying phenomenon, and so we opt to refer to
them by a single empirically descriptive name.  We also considered
names that focused on the presence of transient corotating clumps, or
more generally on circumstellar plasma tori; for the time being,
``CQV'' is agnostic to the true underlying phenomenon.

@@

		\vspace{-0.6cm}
		\subfloat{
			\includegraphics[width=0.8\textwidth]{f1b.pdf}
		}
	\end{center}
		\vspace{-0.3cm}
	\caption{
		{\bf Complex quasiperiodic variables (CQVs)}:
    {\it Top:} Phase-folded TESS light curves of three CQVs.  Each is
    phase-folded over one month.  Gray circles are raw 2-minute data; black
    circles bin to 300 points per cycle.  Periods in hours are in the
    bottom right of each panel.  Left-to-right, the objects are LP
    12-502 (TIC 402980664; Sector~19), TIC 94088626 (Sector 10), and
    TIC 425933644 (Sector~28).
    {\it Bottom:} Cartoon explanations for the phenomenon.  The dust clump
    scenario (lower left) and prominence scenario (lower center) both
    invoke centrifugally-supported material at the corotation radius.
    We disfavor the screen scenario (see Section~\ref{sec:intro}).
	}
	\label{fig:f1}
\end{figure*}

@@

Our search confirms that the CQV phenomenon persists for at least
$\approx$150\,Myr.  Table~\ref{tab:thetable} includes three
$\approx$150\,Myr CQVs in AB~Dor \citep{2015MNRAS.454..593B}, a
$\approx$112\,Myr old Pleiades CQV \citep{2015ApJ...813..108D}, and a
similarly-aged Psc-Eri member \citep{2020A&A...639A..64R}.  To our
knowledge, TIC~332517282 in AB~Dor ($t$$=$$149^{51}_{-19}$\,Myr;
\citealt{2015MNRAS.454..593B}) was the previous record-holder for the
oldest-known CQV \citealt{2019ApJ...876..127Z,2022AJ....163..144G}; at
least one unambiguous CQV (EPIC~211070495) and a few other candidates
were also previously known in the Pleiades
\citep{2016AJ....152..114R}.

@@

At their core, the planetesimals would primarily serve as the source
for the occulting gas or dust; they would not necessarily need to
explain the observed phases of the observed dips.  The azimuth angle
of the eventual confinement could be entirely determined by the
stellar magnetic field configuration.  In this scenario, the obscuring
material could potentially inspiral from outgassing planetesimals many
stellar radii away from the corotation radius.  While the
planetesimals themselves would not necessarily need to transit, if
they did, they would need to be $\lesssim 1$\,$R_\oplus$ based on
their non-detections in the TESS data.  This would be analogous to
previously known systems such as K2-22 \citep{2015ApJ...812..112S} and
KOI-2700 \citep{2014ApJ...784...40R}, though the obscuring material
would be observed much further from the emitting planet.

@@

>>>>

All young stars vary in optical brightness, and the origin of such
variability is, in most cases, understood.  Well-explored sources of
optical variability include inhomogeneities on stellar surfaces such
as starspots and faculae \citep[e.g.][]{2021isma.book.....B},
occultations by circumstellar disks
\citep[e.g.][]{2017MNRAS.470..202B}, and, in geometrically favorable
circumstances, eclipses by stars and planets
\citep[e.g.][]{2020AJ....160...33R}.  More exotic sources of optical
variability that are potentially relevant to this work include
transiting exocomets (e.g.~$\beta$~Pic;
\citealt{2019A&A...625L..13Z}), disintegrating rocky bodies (e.g.
KOI-2700; \citealt{2014ApJ...784...40R}), and occultations by
circumstellar plasma clumps (e.g.~$\sigma$~Ori~E;
\citealt{2005ApJ...630L..81T,2005MNRAS.357..251T}).

Data from K2 \citep{2014PASP..126..398H} and TESS
\citep{2015JATIS...1a4003R} have revealed a new class of variable
star for which the root cause of variability is only beginning to
become clear: complex quasiperiodic variables (CQVs).  These objects
are identified from their optical light curves, which show nearly
periodic troughs that are either sharp or broad; these troughs are
often superposed on quasi-sinusoidal spot-like modulation
\citep{2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z}.
Some CQVs show up to eight dips per cycle.  Most CQVs are
pre-main-sequence M dwarfs with ages of $\approx$5-150 million years
(Myr), and rotation periods of 0.2--2 days.  They are observed to
comprise $\approx$1-3\% of M dwarfs younger than 100 Myr
\citep{2016AJ....152..114R,2022AJ....163..144G}.  They generally do
not show near-infrared excesses indicative of dusty disks, but the
wavelength-dependent dip amplitudes of some CQVs is consistent with
reddening by dust
\citep{2017PASJ...69L...2O,2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
The dip amplitudes and phases usually evolve gradually over tens to
hundreds of cycles, although they have occasionally been observed to
change abruptly within one cycle
\citep[e.g.][]{2017AJ....153..152S,2022ApJ...925...75P,2023ApJ...945..114P}.

The sharp features of CQV light curves can have durations as short as
5\% of the rotation period ($P_{\rm rot}$), which is too short to be
caused by starspots rotating into and out of view.  Starspots produce
flux variations with characteristic timescales of $P_{\rm rot}$ and
$0.5\,P_{\rm rot}$.  With finely-tuned viewing geometries, starspots
can produce dip durations as short as $\approx$$0.2\,P_{\rm rot}$, but
in such cases, limb darkening causes the dip amplitudes to be smaller
than the observed amplitudes of $\sim$1\% (see
\citealt{2017AJ....153..152S}, Figures 37-41).  Thus, a
``starspot-only'' scenario can be ruled out for many CQVs
\citep{2017AJ....153..152S,2019ApJ...876..127Z,2021MNRAS.500.1366K}.
Given that many CQVs cannot be explained by starspots alone, and
working under the assumption that all CQVs share the same basic
physical scenario, we discard the ``starspot-only'' scenario.
Instead, the correct explanation probably involves material above the
stellar surface that is spatially confined with sharp boundaries
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.

Figure~\ref{fig:f1} illustrates two proposed models for the extrinsic
material.  The first scenario invokes opaque dust ``clumps'' that
orbit near the Keplerian corotation radius [$R_{\rm cr} =
(GM/\Omega^2)^{1/3}$, where $\Omega = 2\pi/P_{\rm rot}$] and
periodically transit the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The second scenario
invokes ``prominences'', long-lived condensations of cool, dense,
marginally-ionized gas that are embedded within the hotter corona and
corotate with the star
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
These hypothetical structures are analogous to quiescent prominences
and filaments seen in the solar corona \citep[see
e.g.][]{2015ASSL..415.....V}, though rather than existing at a
fraction of the stellar radius as in the solar case, they would exist
at distances of a few stellar radii.  A final possibility is that a thin
optically-thick ring obscures a small portion of the stellar
photosphere \citep{2019ApJ...876..127Z}.  Hot spots passing behind such a structure could produce
sudden dips.  We do not favor this scenario due to the issues
discussed in Appendix~\ref{app:ring}.

The dust clump and prominence hypotheses are qualitatively similar,
except that one invokes opacity from dust, while the other invokes
opacity from gas, likely bound-free transitions in hydrogen or perhaps
a molecular opacity.  However unambiguous evidence in support of any particular
scenario has yet to be acquired.  Such evidence might include a
spectroscopic detection of silicate 10\,$\mu$m dust absorption during
a dip, or perhaps detection of transient Balmer-line excesses as a
function of cycle phase, similar to observations made in systems such
as AB~Dor \citep[see][]{1999ASPC..158..146C} or PTFO 8-8695
\citep{2016ApJ...830...15J}.

CQVs remain mysterious because they have been both hard to discover
and hard to characterize.   They are hard to discover because they are
rare: CQVs comprise $\approx$1\% of the youngest $\approx$1\% of M
dwarfs \citep{2018AJ....155..196R}.  Out of the millions of stars
monitored by K2 and TESS, about 50 CQVs have been reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2022AJ....163..144G,2023ApJ...945..114P}.
They have been hard to characterize because many of the known CQVs are
relatively faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, yielding optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  At such magnitudes, high-resolution
time-series spectroscopy is out of reach with current facilities,
despite the potential utility of such observations.

In this work, we aim to find bright and nearby CQVs, since these
objects will be the most amenable to detailed photometric and
spectroscopic analyses.  To do this, we use 120-second cadence data
acquired by TESS between 2018 July and 2022 September (Sectors 1-55;
Cycles 1-4).  We present our search methods in
Section~\ref{sec:methods}, and the resulting CQV catalog in
Section~\ref{sec:results}.  The observed evolution of many CQVs over a
two-year baseline is described in Section~\ref{sec:evoln}, including a
deep-dive into the behavior of an especially interesting object, LP
12-502.  We discuss a few implications in
Section~\ref{sec:discussion}, and conclude in
Section~\ref{sec:conclusion}.

Some comments on nomenclature are needed.  What we are calling CQVs
have also been called ``transient flux dips'', ``persistent flux
dips'', ``scallop shells'', ``batwings'', \citep{2017AJ....153..152S}
``complex rotators'',
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P},
and ``complex periodic variables'' \citep{2023MNRAS.518.2921K}.  The
CQVs should not be conflated with ``dippers'', which are classical T
Tauri stars with infrared excesses, and which show large-amplitude
variability linked to obscuring inner disk structures and accretion
hot spots \citep{2014AJ....147...82C,2021ApJ...908...16R}.  The
phenomenology and stellar properties of CQVs and dippers are
quite different (though see Sections~\ref{subsec:irexcess}
and~\ref{subsec:discdippers}).  Despite the cost of introducing yet
another naming standard, we prefer to use a single empirically
descriptive name that highlights these objects as variable stars, and
not ``rotators''.  We also prefer a nomenclature that reflects how the
dips and peaks of these stars always appear to be transient 
when observed over timescales exceeding hundreds of cycles.  ``Complex
quasiperiodic variable'' meets these requirements, for the time being.

@@

		\vspace{-0.6cm}
		\subfloat{
			\includegraphics[width=0.66\textwidth]{f1b.pdf}
		}
	\end{center}
		\vspace{-0.3cm}
	\caption{
		{\bf Complex quasiperiodic variables (CQVs)}:
    {\it Top:} Phase-folded TESS light curves of three CQVs.  Each panel shows
    the average of the data accumulated over one month.
    Gray circles are raw 2-minute data; black
    circles are binned to 300 points per cycle.  The periods in hours
    is printed in the
    bottom right corner.  Left-to-right, the objects are LP
    12-502 (TIC 402980664; Sector~19), TIC 94088626 (Sector 10), and
    TIC 425933644 (Sector~28).
    {\it Bottom:} Cartoon explanations for the phenomenon.  The dust
    clump scenario (left) and prominence scenario (right) both invoke
    magnetically-confined corotating material.
	}
	\label{fig:f1}
\end{figure*}

@@

Our search confirms that the CQV phenomenon persists for at least
$\approx$150\,Myr.  Table~\ref{tab:thetable} includes three
$\approx$150\,Myr CQVs in AB~Dor \citep{2015MNRAS.454..593B}, a
$\approx$112\,Myr old Pleiades CQV \citep{2015ApJ...813..108D}, and a
similarly-aged Psc-Eri member \citep{2020A&A...639A..64R}.  To our
knowledge, TIC~332517282 in AB~Dor ($t$$=$$149^{51}_{-19}$\,Myr;
\citealt{2015MNRAS.454..593B}) was the previous record-holder for the
oldest-known CQV \citep{2019ApJ...876..127Z,2022AJ....163..144G}; at
least one unambiguous CQV (EPIC~211070495) and a few other candidates
were also previously known in the Pleiades
\citep{2016AJ....152..114R}.

@@

The main logical function of the planetesimals is to serve as a source
for the occulting gas or dust; they would not necessarily need to
explain the observed phases of the observed dips.  The azimuth angle
of the eventual confinement could be entirely determined by the
stellar magnetic field configuration.  In this scenario, the obscuring
material could potentially inspiral from outgassing planetesimals many
stellar radii away from the corotation radius.  While the
planetesimals themselves would not necessarily need to transit, if
they did, they would need to be $\lesssim 1$\,$R_\oplus$ based on
their non-detections in the TESS data.  This would be analogous to
previously known systems such as K2-22 \citep{2015ApJ...812..112S} and
KOI-2700 \citep{2014ApJ...784...40R}, though the obscuring material
would be observed much further from the emitting planet.

@@

\section{The ring hypothesis}
\label{app:ring}

One hypothesis for the CQVs, presented by \citet{2019ApJ...876..127Z},
is that the star might be ``{\it orbited by one or more rings composed of
dust-size or somewhat larger particles\ldots The ring particles would
move in Keplerian orbits at relatively large distances from the star,
and therefore the sublimation lifetime would not be an issue even if
the particles are dust-like in size.}'' A sketch of this scenario is
presented by \citet{2019ApJ...876..127Z}, in their Figure~11.
An example set of proposed parameters involved a ring inclined with
respect to the stellar spin axis by a few degrees, and with inner and
outer radii of 10 and 15 stellar radii.

One concern with the ring hypothesis is that if a cool spot were to
transit behind the ring, it would produce a brightening, rather than a
dimming.   Most CQVs show dimmings.   The ring scenario would
therefore imply that large hot spots are common in the photospheres of
pre-main-sequence M dwarfs.  Empirical evidence however suggests that
the optical variability of disk-free pre-main-sequence stars is
dominated by cool spots.  This evidence includes flux excursions
caused by spot-crossings during planetary transits
\citep[e.g.][]{2020AJ....160...33R,2022AJ....163..147G}, correlations
between simultaneous photometric and chromospheric time-series
\citep{2019A&A...621A..21R}, and stellar spectra that show molecules
that only form at cool temperatures
\citep[e.g.][]{2017ApJ...836..200G,2023ApJ...946...10P}.

A stronger concern with the ring hypothesis is that it is fine-tuned.
The model requires specific locations for the inner edge and the outer
edge of the ring, an inclination that yields a band with a particular
apparent size, and for material in the ring to be optically thick
while also being homogeneous enough to not induce any apparent
photometric variability.
While it is challenging to ascribe specific probabilities to any one
of these factors, 
the requirement that they all be simultaneously met seems sufficiently
severe to disfavor this scenario.
